Inferring community characteristics in labelled networks

### 2021-05-19 nips/appendices.tex
Much work
First draft

>>>>

\subsection{Inferring the number of blocks}

For the purposes of our model (the FFBM), the number of blocks $B$ is a constant which must be specified by the data scientist. We could however, allow our choice of $B$ to be influenced by the observed data. This places us in the domain of empirical Bayes, which must be negotiated carefully. Prior beliefs must be determined a priori else they are not prior. However, as the number of blocks only specifies the coarseness of the analysis, it is fine to allow it to vary. Indeed, \citet{peixoto-determine-B} shows that for a fixed average degree the maximum number of detectable blocks scales as $O(\sqrt{N})$ where $N$ is the number of vertices.

If we allow $B$ to vary in the $b$-chain (i.e. new blocks can be created and we permit empty blocks) then it can be run  until a minimum description length (MDL) solution is reached. We take the number of non-empty blocks at the MDL to be our fixed block number $B$ for subsequent analysis. Indeed, it is prudent to start our $b$-chain at this MDL solution as then the burn-in time is greatly reduced.

\subsection{Implementation details}
\label{sec:imp-details}

All data analysis and visualisation was implemented in Python. Full source code is available at:

\url{https://github.com/LozzaTray/Jormungandr}

The scripts were run using a standard PC. Specs are:

\begin{itemize}
	\item \textbf{CPU}: Intel(R) Core(TM) i7-1065G7
	\item \textbf{RAM}: 8GB
	\item \textbf{GPU}: Intel(R) Iris(R) Plus Graphics
\end{itemize}

### 2021-05-19 nips/conclusion.tex
First draft

The FFBM was developed to address the shortcomings of other graphical models when trying to test how vertex features affect graphical. The idea is to divide the graph into its most natural partition and only then bring in the vertex features to determine which ones best predict the partition. The inference algorithm proposed sidesteps an expensive summation over all latent block-states through the use of the microcanonical rather than the traditional SBM

The overall method is shown to be effective at extracting and describing the most natural communities in a labelled network. Nevertheless, the approach can only currently explain the structure on the macro-scale. Future work may benefit from extending the FFBM to be hierarchical in nature. That way, the structure of the network can be explained at each length-scale.

>>>>

The FFBM was developed to address the shortcomings of other graphical models when testing how vertex features affect community structure. The idea is to divide the graph into its most natural partition and only then bring in the vertex features. We therefore hope to avoid a common pitfall in data science: correlation does not imply causation. It is very easy to find vertex features that are in some way correlated with the graphical structure. Nonetheless, only when we find the feature that best describes the most pronounced partition do we have a stronger case for causation.

With the newly-defined FFBM, we go on to present an efficient inference algorithm to ascertain the parameters $\theta$ of the block generator. This is introduced as two concurrent Markov chains to sample the block memberships $b$ and block generator parameters $\theta$. The key trick to make the $b$-chain efficient is the the use of the microcanonical rather than the traditional SBM (as proposed by \citet{Peixoto-Bayesian-Microcanonical}) which allows us to bypass an expensive summation over the latent memberships $b$. Additionally, we use the empirical mean of the $b$-samples as the input to our $\theta$-chain, as this reduces the variance in our evaluation of the target distribution and thus shortens the burn-in and auto-correlation times of our $\theta$-chain.

The overall method is shown to be effective at extracting and describing the most natural communities in a labelled network. Nevertheless, the approach can only currently explain the structure at the macro-scale. We cannot explain structure within each community. Future work will benefit from extending the FFBM to be hierarchical in nature. That way, the structure of the network can be explained at all length-scales of interest.

### 2021-05-19 nips/experiments.tex
First draft

\begin{itemize}
	\item \textbf{Political books} \cite{polbooks} ($N=105, E=441, D=1$) - network of Amazon book sales about U.S. politics, published close to the presidential election in 2004. Two books are connected if they were frequently co-purchased by the same customer. Vertex features encode the broad political affiliation of the author (liberal, conservative, neutral).

	\item \textbf{Primary school dynamic contacts} \cite{schools} ($N=238, E=5539, D=13$) - network of face-to-face contacts amongst students and teachers at a primary school in Lyon, France. Two nodes are connected if the two parties shared a face-to-face interaction over the course of the day. Vertex features include class membership, gender and whether or not the student is a teacher or pupil. These data were collected on consecutive days in October 2009. We choose to analyse just the second day.

@@

\begin{table}[!h]
	\centering
	\caption{FFBM fit for the datasets}
	\label{tab:results}
	\begin{tabular}{c|ccc|c|cc}
		Dataset & $N$ & $E$ & $D$ & $B$ & $\bar{S}_b /N$ & $\bar{U}_\theta /N$ \\ \hline
		Political books & 105  & 441 & 1 & 3 & 11.7 & 0.707 \\
		Primary school & 238 & 5539 & 13 &  18 & 43.0  & 2.75 \\
		FB egonet & 747 & 30025 & 480 & 10  &           &             \\

	\end{tabular}
\end{table}

\subsection{Political books}

This dataset was collected by \citet{polbooks}. We wish to determine whether political affiliation is a good predictor of the overall network structure. We choose to partition the network into $B=3$ communities as we only have this many distinct values for political affiliation (conservative, liberal or neutral). The inferred block memberships are given in figure \ref{fig:books-graph}.

\begin{figure}[!h]
	\centering
	\begin{subfigure}{0.3\linewidth}
		\centering
		\includegraphics[width=\linewidth]{polbooks-graph.png}
		\fbox{\includegraphics[width=0.4\linewidth]{3-legend.png}}
		\caption{Graph coloured by inferred block memberships of each node}
		\label{fig:books-graph}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.5\linewidth}
		\centering
		\includegraphics[width=\linewidth]{polbooks-null.png}
		\caption{Feature-to-block generator sampled weight parameters for each block index}
		\label{fig:book-null}
	\end{subfigure}
	\caption{Political books network}
\end{figure}

\subsection{Primary school dynamic contacts}

These data were originally collected by \citet{schools} to quantify the transmission opportunities for respiratory infections within a primary school context (ages 6-11 in France). However, we seek to ask a simpler question. What features best describe how people interact with one another in a primary school context. The only vertex features we have available are school-class (one of 10 values - 2 per year group), gender and a distinction between teachers and pupils.

We must first choose the number of blocks to consider $B$ to define the coarseness of our analysis. A total of 10 classes would suggest that $B=10$ is a natural starting point. With this value of $B$ we first sample from the block membership posterior $b^{(t)} \sim p(b | A, X)$. We visualise the inferred partition in figure \ref{fig:school-graph}

\begin{figure}[!h]
	\centering
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{school-graph.png}
		\fbox{\includegraphics[width=0.5\linewidth]{school-legend.png}}
		\caption{Graph coloured by inferred block memberships of each node}
		\label{fig:school-graph}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{school-null.png}
		\caption{Feature-to-block generator sampled weight parameters for each block index}
		\label{fig:school-null}
	\end{subfigure}
	\caption{Primary school dynamic contacts network}
\end{figure}

@@

\begin{figure}[!h]
	\centering
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{fb-graph.png}
		\fbox{\includegraphics[width=0.3\linewidth]{10-legend.png}}
		\caption{Graph coloured by inferred block memberships of each node}
		\label{fig:fb-graph}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{fb-null.png}
		\caption{Feature-to-block generator sampled weight parameters for each block index}
		\label{fig:fb-null}
	\end{subfigure}
	\caption{Primary school dynamic contacts network}
\end{figure}

>>>>

\begin{itemize}
	\item \textbf{Political books} \cite{polbooks} ($N=105, E=441, D=3$) - network of Amazon book sales about U.S. politics, published close to the presidential election in 2004. Two books are connected if they were frequently co-purchased by the same customer. Vertex features encode the broad political affiliation of the author (liberal, conservative or neutral).

	\item \textbf{Primary school dynamic contacts} \cite{schools} ($N=238, E=5539, D=13$) - network of face-to-face contacts amongst students and teachers at a primary school in Lyon, France. Two nodes are connected if the two parties shared a face-to-face interaction over the course of the day. Vertex features include class membership, gender and whether or not the individual is a teacher or pupil. These data were collected on consecutive days in October 2009. We choose to analyse just the second day.

@@

\begin{table}[!h]
	\centering
	\caption{FFBM fit for the datasets}
	\label{tab:results}
	\begin{tabular}{c|ccc|c|cc}
		Dataset & $N$ & $E$ & $D$ & $B$ & $\bar{S}_b /N$ & $\bar{U}_\theta /N$ \\ \hline
		Political books & 105  & 441 & 3 & 3 & 11.7 & 0.707 \\
		Primary school & 238 & 5539 & 13 &  10 & 45.9  & 1.61 \\
		FB egonet & 747 & 30025 & 480 & 10  & 66.9 & 7.47
	\end{tabular}
\end{table}

\FloatBarrier
\subsection{Political books}

This dataset was collected by \citet{polbooks}. We wish to determine whether political affiliation is a good predictor of the overall network structure. We choose to partition the network into $B=3$ communities as we only have this many distinct values for political affiliation (conservative, liberal or neutral). The inferred block memberships are given in figure \ref{fig:books-graph}. We sample the block-generator parameters and plot the emprical mean and standard deviation of each sample on figure \ref{fig:book-null}.
%
\begin{figure}[!h]
	\centering
	\begin{subfigure}{0.3\linewidth}
		\centering
		\includegraphics[width=\linewidth]{polbooks-graph.png}
		\fbox{\includegraphics[width=0.4\linewidth]{3-legend.png}}
		\caption{Sampled block memberships $\hat{Y}$}
		\label{fig:books-graph}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.5\linewidth}
		\centering
		\includegraphics[width=\linewidth]{polbooks-null.png}
		\caption{Feature weights}
		\label{fig:book-null}
	\end{subfigure}
	\caption{Political books network \cite{polbooks}}
\end{figure}
%
Indeed, for all 3 blocks, each has a distinct political affiliation as its highest magnitude component. This is strong evidence that political affiliation is indeed the axis which best predicts the 3-way natural partition of the graph into blocks. Nevertheless, for block index 1, although \verb*|neutral| is the best predictor of the ones we have available, the value is not particularly extreme. Perhaps, there is an unobserved feature that places a book into this block. Indeed, block-0 is more of a centre-right block than true centre.

\FloatBarrier
\subsection{Primary school dynamic contacts}

These data were originally collected by \citet{schools} to quantify the transmission opportunities for respiratory infections. However, we seek to ask which vertex features best describe how people interact with one another in a primary school context. The only vertex features we have available are school-class (one of 10 values - 2 per year group), gender and a distinction between teachers and pupils.

We must first choose the number of blocks $B$ to define the coarseness of our analysis. A total of 10 school-classes would suggest that $B=10$ is a natural starting point. We visualise the inferred block memberships in figure \ref{fig:school-graph}.
%
\begin{figure}[!h]
	\centering
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{school-graph.png}
		\fbox{\includegraphics[width=0.5\linewidth]{10-legend.png}}
		\caption{Sampled block memberships $\hat{Y}$}
		\label{fig:school-graph}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{school-null.png}
		\caption{Reduced dimension feature weights}
		\label{fig:school-null}
	\end{subfigure}
	\caption{Primary school dynamic contacts network \cite{schools}}
\end{figure}

As before, we proceed to sample the block-generator parameters $\theta$ and employ the dimensionality reduction heuristic defined in section \ref{sec:dim-reduction} with null threshold $c=1$ and standard deviation multiplier $k=1$. We then plot the weights for the features that survive the discard step in figure \ref{sec:dim-reduction}. Immediately, we see that only the pupils' class memberships have survived (1A-5B); gender and teacher/student status have been discarded meaning that these are not good predictors of overall macro-structure.

The vast majority of blocks are composed of a single class. However, some blocks have 2 comparably good classes as their predictor. For example, blocks 2 and 4 contain classes 4A and 4B as their 2 best predictors. This suggests that the social divide between classes is less pronounced for pupils in year 4 but there is nonetheless an unobserved variable that splits year 4 into 2 communities. The most surprising block is number 5 - which has comparable weightings for classes 5A and 1B. Perhaps there was a joint event between those two classes on the day the data were collected.

@@

This dataset was chosen to showcase the power of the dimensionality reduction technique as the feature-space has high dimension ($D=480$). We sample the $b$-chain specifying $B=10$ total blocks and use this to construct the $\theta$-samples as before.

We apply the dimensionality reduction heuristic outlined in section \ref{sec:dim-reduction}, choosing $k=1$ and $c=0.8$. This leads to a reduced dimension $D'=13$. These parameter values are plotted on figure \ref{fig:fb-null}. The features that remain are those that best explain the high-level community structure. More so than language, gender or surname - the school\footnote{This includes higher education institutes} you attend is the best explanatory variable for how the observed network's communities are partitioned at the highest level. 
%
\begin{figure}[!h]
	\centering
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{fb-graph.png}
		\fbox{\includegraphics[width=0.3\linewidth]{10-legend.png}}
		\caption{Sampled block memberships $\hat{Y}$}
		\label{fig:fb-graph}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{fb-null.png}
		\caption{Reduced dimension feature weights}
		\label{fig:fb-null}
	\end{subfigure}
	\caption{Facebook egonet id 1912 \cite{fb-snap}}
\end{figure}

Nevertheless, this example also highlights a weakness of the method for high-dimensional feature spaces. When the feature dimension is very large, it becomes increasingly likely that a particular feature may uniquely identify a small set of nodes. If these nodes are all part of the same community then the classifier will overfit for that particular parameter. The regularisation term imposed by the prior goes some way to alleviating this problem but we see in figure \ref{fig:fb-null} that the feature \verb*|birthday-5| has a very high weight as it relates to block 0. It might be possible to shift the feature values such that they take values in $\{-1, 1\}$ rather than $\{0, 1\}$ but this approach falls down when input features are mutually correlated. Indeed, for features with discrete values such as the class memberships (1A, 1B, 2A \dots 5B) it is preferred to keep $x_i \in \{0, 1\}^D$ as then each block can be readily accept more than one mutually exclusive feature-group (i.e. two school-classes are part of the same block).

### 2021-05-19 nips/inference.tex
First draft

\section{Inference}

@@

\subsection{Choosing the number of blocks}

For the purposes of our model (the FFBM), the number of blocks $B$ is a constant which must be specified by the data scientist. We could however, allow our choice of $B$ to be influenced by the observed data. This places us in the domain of empirical Bayes, which must be negotiated carefully. Prior beliefs must be determined a priori else they are not prior. However, as the number of blocks only specifies the coarseness of the analysis, it is fine to allow it to vary. Indeed, \citet{peixoto-determine-B} shows that for a fixed average degree the maximum number of detectable blocks scales as $O(\sqrt{N})$ where $N$ is the number of vertices.

If we allow $B$ to vary in the $b$-chain (i.e. new blocks can be created and we permit empty blocks) then it can be run  until a minimum description length (MDL) solution is reached. We take the number of non-empty blocks at the MDL to be our fixed block number $B$ for subsequent analysis. Indeed, it is prudent to start our $b$-chain at this MDL solution as then the burn-in time is greatly reduced.

\subsection{Dimensionality reduction}

Once we have the samples $\left\{ \theta^{(t)} \right\}_{t=1}^{T_\theta} \sim p(\theta | A, X)$, we can compute the empirical mean and standard deviation of each component of $\theta$. Switching back to matrix notation we define $\theta = W$, such that $W_{ij}$ is the weight component for block $i$ and feature $j$, we can define:
%
\begin{equation}
	\hat{\mu}_{ij} \coloneqq \frac{1}{T_\theta} \sum_{t=1}^{T_\theta} W_{ij}^{(t)} \qquad \textrm{and} \qquad
	\hat{\sigma_{ij}} \coloneqq \frac{1}{T_\theta} \sum_{t=1}^{T_\theta} \left( W_{ij}^{(t)} - \hat{\mu}_{ij} \right)^2
\end{equation}
%
A simple heuristic to discard the least important features requires specifying a cutoff $c > 0$ and a multiplier $k > 0$. We then discard all features $j$ wuch that equation \ref{eqn:discard} is satisfied.
%
\begin{equation}
	(\hat{\mu}_{ij} - k \hat{\sigma}_{ij}, \hat{\mu}_{ij} + k \hat{\sigma}_{ij}) \cap (-c, +c) = \emptyset \quad \forall i \in \{1,2 \dots B\}
	\label{eqn:discard}
\end{equation}
%
Intuitively, this means discarding any feature for which $\hat{\mu}_{ij} \pm k\hat{\sigma_{ij}}$ lies within or spans the null region $(-c, c)$. If we were to use the Laplace approximation for the posterior $p(W_{ij} | A, X) \approx \Gaussian(W_{ij}; \mu_{ij}, \sigma_{ij})$, then this is effectively a hypothesis test on the value of $W_{ij}$.
%
\begin{equation}
	\begin{aligned}
		H_0&: |W_{ij}| \leq c \\
		H_1&: |W_{ij}| > c
	\end{aligned}
\end{equation}
%
With the multiplier $k$ specifying the degree of significance of the result (equivalent to the probability of a type-1 error). However, as the approximation is not exact we will only treat this dimensionality reduction method as a useful heuristic.

>>>>

\section{Inference}
\label{sec:inference}

@@

\subsection{Dimensionality reduction}
\label{sec:dim-reduction}

Once we have the samples $\left\{ \theta^{(t)} \right\}_{t=1}^{T_\theta} \sim p(\theta | A, X)$, we can compute the empirical mean and standard deviation of each component of $\theta$. Switching back to matrix notation we define $\theta = W$, such that $W_{ij}$ is the weight component for block $i$ and feature $j$, we can define:
%
\begin{equation}
	\hat{\mu}_{ij} \coloneqq \frac{1}{T_\theta} \sum_{t=1}^{T_\theta} W_{ij}^{(t)} \qquad \textrm{and} \qquad
	\hat{\sigma_{ij}} \coloneqq \frac{1}{T_\theta} \sum_{t=1}^{T_\theta} \left( W_{ij}^{(t)} - \hat{\mu}_{ij} \right)^2
\end{equation}
%
A simple heuristic to discard the least important features requires specifying a cutoff $c > 0$ and a multiplier $k > 0$. We then discard all features $j$ wuch that equation \ref{eqn:discard} is satisfied.
%
\begin{equation}
	(\hat{\mu}_{ij} - k \hat{\sigma}_{ij}, \hat{\mu}_{ij} + k \hat{\sigma}_{ij}) \cap (-c, +c) = \emptyset \quad \forall i \in \{1,2 \dots B\}
	\label{eqn:discard}
\end{equation}
%
Intuitively, this means discarding any feature for which $\hat{\mu}_{ij} \pm k\hat{\sigma_{ij}}$ lies within or spans the null region $(-c, c)$. If we were to use the Laplace approximation for the posterior $p(W_{ij} | A, X) \approx \Gaussian(W_{ij}; \mu_{ij}, \sigma_{ij})$, then this is effectively a hypothesis test on the value of $W_{ij}$.
%
\begin{equation}
	\begin{aligned}
		H_0&: |W_{ij}| \leq c \\
		H_1&: |W_{ij}| > c
	\end{aligned}
\end{equation}
%
With the multiplier $k$ specifying the degree of significance of the result (equivalent to the probability of a type-1 error). However, as the approximation is not exact we will only treat this dimensionality reduction method as a useful heuristic and not an exact method.

### 2021-05-19 nips/prelim.tex
First draft

This section defines some preliminary concepts required for the subsequent analysis. We first need a model for community-like structure in a graphical network. For this we adopt the stochastic block model (SBM) - widely used across academia. The premise is that each node in the graph belongs to a unique community called a block. The probability that two nodes are connected depends only on the block memberships of each. Graphs drawn from this ensemble exhibit community structure. Specifically, we will use the microcanonical variant of the SBM, proposed by \citet{Peixoto-Bayesian-Microcanonical}. A paraphrased definition is given below.

@@

Nevertheless, this formulation does not accept high degree variability within blocks as is typical of real-world data. Indeed, the NDC-SBM favours a partition into high-degree and low-degree nodes rather than clusters of inter-connected nodes. We therefore introduce the degree-corrected SBM (DC-SBM) to circumvent these issues.

>>>>

This section defines some preliminary concepts required for the subsequent analysis. We first need a model for community-like structure in a graphical network. For this we adopt the stochastic block model (SBM) - widely used across academia. The premise is that each node in the graph belongs to a unique community called a block. The probability that two nodes are connected depends only on the block memberships of each. Graphs drawn from the SBM ensemble exhibit community structure. Specifically, we will use the microcanonical variant of the SBM, proposed by \citet{Peixoto-Bayesian-Microcanonical}. A paraphrased definition is given below for the non-degree-corrected SBM (NDC-SBM).

@@

Nevertheless, this formulation does not accept high degree variability within blocks as is typical of real-world data. Indeed, the NDC-SBM favours a partition into high-degree and low-degree nodes rather than clusters of inter-connected nodes. We therefore introduce the degree-corrected SBM (DC-SBM) \cite{Peixoto-Bayesian-Microcanonical} to circumvent these issues.
